---
title: "ps5-markdown/basic-plots"
author: "Anna Sun"
date: "2023-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

# 1 Load and check data (5pt)

You first task is to do a very simple data check:

1.  (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See [**https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating**](https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating).\

done the working directory for this new rmarkdown document is `r getwd()`

```{r message=FALSE}
library(tidyverse)
```

2.  (2pt) Load data. How many rows/columns do we have?\
```{r}
data <- read_delim("gapminder.csv")
```

there are `r nrow(data)` rows and `r ncol(data)` columns in the data

3.  (2pt) Print a small sample of data. Does it look OK?
```{r}
data %>% 
  sample_n(5)
```
yes the data looks okay

------------------------------------------------------------------------

# 2 Descriptive statistics (15pt)

1.  (3pt) How many countries are there in the dataset? Analyze all three: *iso3*, *iso2* and *name*.
```{r}
cat("iso3:", length(unique(data$iso3)), "\n")
cat("iso2:", length(unique(data$iso2)), "\n")
cat("name:", length(unique(data$name)), "\n")
```


2.  If you did this correctly, you saw that there are more iso-2 codes than names, and there are even more *iso3*-codes. What is going on? Can you find it out?

    (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
```{r}
data %>% 
  group_by(iso2) %>% 
  summarize(num_of_names = n_distinct(name)) %>% 
  arrange(desc(num_of_names))
```
?? there is one more name than iso2 codes, this is because 2 names do not have a iso2 code

    (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are *CHANISL* and *NLD CURACAO*.
```{r}
data %>% 
  group_by(name) %>% 
  summarize(num_of_iso3 = n_distinct(iso3)) %>% 
  arrange(desc(num_of_iso3))
```
?? confused if i did this wrong because it did not match up with the hint, I got that 4 iso3 codes did not match up with a name though

3.  (2pt) What is the minimum and maximum year in these data?
```{r}
times <- data %>% 
  select(time) %>% 
  filter(!is.na(time))

max(times)
min(times)
```

------------------------------------------------------------------------

# 3 CO2 emissions (30pt)

Next, let's analyze CO2 emissions.

1.  (2pt) How many missing co2 emissions are there for each year? Analyze both missing *CO2* and *co2_PC*. Which years have most missing data?\
```{r}
data %>% 
  select(time, co2, co2_PC) %>% 
  filter(!is.na(time)) %>% 
  group_by(time) %>% 
  summarize(missing_co2 = sum(is.na(co2)), missing_co2_PC = sum(is.na(co2_PC))) %>% 
  arrange(desc(missing_co2))
```
based on the data 2017, 2018, and 2019 had the most missing data

2.  (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.\
```{r}
co2_overtime <- data %>% 
  filter(iso3 %in% c("USA", "CHN", "IND", "AGO", "AUS"))

ggplot(co2_overtime, aes(time, co2, col=iso3)) +
  geom_line()
```
china over time has had the greatest growth in total co2 emissions, while angola/australia haven't changed much over the years

3.  (5pt) Now let's analyze the CO2 emissions per capita (*co2_PC*). Make a similar plot of the same countries. What does this figure suggest?\
```{r}
ggplot(co2_overtime, aes(time, co2_PC, col=iso3)) +
  geom_line()
```
this plot shows that the USA actually had the most co2 emission per capita and though china had the greatest total release has a much lower co2 emission per capita

4.  (6pt) Compute average CO2 emissions per capita across the continents (assume *region* is the same as continent). Comment what do you see.\
    Note: just compute averages over countries and ignore the fact that countries are of different size.\
    Hint: Americas 2016 should be 4.80.\
```{r}
data_by_co2 <- data %>% 
  filter(!is.na(co2_PC), !is.na(region), time %in% c(1960, 2016)) %>% 
  group_by(time, region) %>% 
  summarize(average_co2_PC = mean(co2_PC))
data_by_co2
```
Europe/Asia has the largest average co2 per capita while Africa has the lowest, lots of data was not associated with a specific region in 2016

5.  (7pt) Make a barplot where you show the previous results--average CO2 emissions per capita across continents in 1960 and 2016.\
    Hint: it should look something along these lines(**Please refer to the spec for the graph**)\
```{r}
data_by_co2 %>% 
  ggplot(aes(region, average_co2_PC, fill=factor(time))) + 
  geom_col(position = "dodge") +
  labs(title="1960 vs 2016 co2 emission per capita", fill="year" )
```

6.  Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume *region* is continent). (CHANGED YEAR TO 2016 BECAUSE OF MISSING DATA)
looking at the graph from the previous question -
largest CO2 emitters (2016): Europe, Asia, Americas
lowest CO2 emitters (2016): Africa, Oceania, Americas

------------------------------------------------------------------------

# 4 GDP per capita (50pt)

Let's look at GDP per capita (*GDP_PC*)

1.  (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.\
    Comment what do you see there.\
```{r}
data %>% 
  filter(time == 1960) %>% 
  ggplot(aes(GDP_PC, lifeExpectancy, col=region, size=totalPopulation)) +
  geom_point()
```
Countries with higher GDP per capita seem to have a higher life expectancy, which mostly belonging to Europe or the Americas

2.  (4pt) Make a similar plot, but this time use 2019 data only.\
```{r}
data %>% 
  filter(time == 2019) %>% 
  ggplot(aes(GDP_PC, lifeExpectancy, col=region, size=totalPopulation)) +
  geom_point()
```
Africa at the lowest GDP and life expectancy and Europe at the highest

3.  (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years?\

Life expectancy and GDP per capita have grown over the last 60 year but the plot shows a similar trend to the earlier graph but each region has a more distinct GDP per capita and life expectancy

4.  (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?\
    Note: here as *average*I mean just average over countries, ignore the fact that countries are of different size.\
```{r}
data_by_le <- data %>% 
  filter(!is.na(lifeExpectancy), !is.na(region), time %in% c(1960, 2019)) %>% 
  group_by(region, time) %>% 
  summarize(average_le = mean(lifeExpectancy))
data_by_le
```
yes the results match the graph which showed Europe as having the highest life expectancy in both 1960 and 2019, and Africa as having the lowest in both years

5.  (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.\
    Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See [**https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute**](https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute).\
```{r}
le_growth <- data_by_le %>% 
  mutate(growth = average_le - lag(average_le)) %>% 
  select(region, growth) %>% 
  filter(!is.na(growth)) %>% 
  arrange(growth)
le_growth

le_growth %>% 
  ggplot(aes(region, growth)) +
  geom_col()
```
Asia and Africa had the greatest and similar growth in life expectancy, while Europe had the least growth 

6.  (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!\
```{r}
data %>% 
  filter(!is.na(GDP_PC), time %in% c(1960, 2019)) %>% 
  ggplot(aes(GDP_PC, fill=factor(time))) +
  geom_histogram()
```
2019 has much more differing GDP per capita compared to 1960 

7.  (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)\
    Hint: check out the function rank()!\
    Hint2: 17 for 1960.\
```{r}
le_ranking_1960 <- data %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), time == 1960) %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  select(name, rank)

le_ranking_1960$rank[le_ranking_1960$name == "United States of America"]
```

```{r}
le_ranking_2019 <- data %>% 
  filter(!is.na(lifeExpectancy), !is.na(name), time == 2019) %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  select(name, rank)

le_ranking_2019$rank[le_ranking_2019$name == "United States of America"]
```

8.  (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019--what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year?\
    Hint: 0.0904 for 1960.\
```{r}
le_ranking_1960$rank[le_ranking_1960$name == "United States of America"] / nrow(le_ranking_1960)
le_ranking_2019$rank[le_ranking_2019$name == "United States of America"] / nrow(le_ranking_2019)
```
USA is in the top 9% of life expectancy in 1960 and is in the top 23% in 2019
------------------------------------------------------------------------

Finally tell us how many hours did you spend on this PS.
I spent arount 5-6 hours on this ps assignment